### 2024-12-24 jsa2024_rebuttal.md

## Reviewer #1
### Reduction of CEFA - would something as simulation reduction work too? It might be better than determinization and minimization. Does determinization sometimes blow up?

We have not tried something like simulation minimization, but it is interesting advice.

Yes, the determinization sometimes blow up because the size of the  result CEFA is exponential at worst case.

### Thirdly, I did not follow why the form (0,…,0,1,0,…,0), as for the union of two CEFAs, two registers as well as two kinds of transitions (\overrightarrow{v1},\overrightarrow{0},1,0) and (\overrightarrow{0},\overrightarrow{v2},0,1) are introduced.

This is mainly because the accepting condition is global. Assuming that we do not introduce new registers and transitions, we can not decide which accepting condition to use when arriving at a final state. We have added some notes at page 13.

### The Overapproximation on page 22. I am wandering about the transition from an original final state to qs that increments the new counter, and the subsequent self loop on qs. Does this also cause overapproximation? All extensions of the originally accepted words are accepted? If yes, it would deserve a comment.

No, this causes the autmoton reject some word supposing to be accepted. For example, the word

### 2024-12-24 2024-5-13/sections/algorithms.tex
complete complexity analysis

\begin{definition}[$\cefadec$]
Let $x_1, \cdots, x_n$ be string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ be nonempty sets of CEFA over the alphabet $\Sigma$ with $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$ where the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{1, l_1}}, \cdots, R_{\aut_{n, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, moreover, let $\varphi$ be an LIA formula whose free variables are from $ \bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$. Then  the CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are said to be nonempty w.r.t. $\varphi$ if there are assignments $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{i,j}}$
such that $(\theta(x_i); \myvec{v_{i,j}}) \in \Lang(\aut_{i, j})$ and  $\varphi[(\myvec{v_{i,j}}/R_{\aut_{i,j}})]$ is true, for every $i \in [n], j \in [l_i]$.
\end{definition}

@@

At first, we show how to construct a CEFA from a regex where counting operators may occur. Our approach is based on Thompson's construction~\cite{Thompson1968} for regexes without counting operators. Let us start with register-representable regexes defined in the sequel.

@@

Let $e$ be a regex over $\Sigma$. Then an occurrence of counting operators in $e$, say $(e')^{\{m,n\}}$ (or $(e')^{\{m, \infty\}}$), is said to be \emph{register-representable} if $(e')^{\{m,n\}}$ (or $(e')^{\{m, \infty\}}$) is not in the scope of a Kleene star, another counting operator, complement, or language difference in $e$. We say that $e$ is \emph{register-representable} if all occurrences of counting operators in $e$ are register-representable. For instance, $a^{\{2, 6\}} \cap a^{\{4, \infty\}}$ is register-representable, while $\overline{a^{\{2, 6\}}}$ and $(a^{\{2, 6\}})^{\{4,\infty\}}$ are not since $a^{\{2,6\}}$ is in the scope of complement and the counter operator $\{2, \infty\}$ respectively.

@@

For non-register-representable regexes, we first transform them into register-representable regexes by unfolding all the non-register-representable occurrences of counting operators. After that, we utilize the aforementioned procedure to construct CEFA. For instance, $(a^{\{2, 6\}} \concat b^*)^{\{2,\infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa+aaaa) \concat b^*)^{\{2, \infty\}}$. 
%$(a^{\{2, 6\}} \concat b^*) \setminus (a + b)^{\{4, \infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa + aaaa) \concat b^*) \setminus (a+b)(a+b)(a+b)(a+b)(a+b)^*$.

@@

In this section, we present a procedure to solve the $\cefadec$ problem: Suppose that $x_1, \cdots, x_n$ are mutually distinct string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty sets of CEFA over the same alphabet $\Sigma$ where $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$. Moreover, the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, and $\varphi$ is a LIA formula whose free variables are from $\bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$. 
The procedure comprises three steps.

@@

In the following steps, we reduce the non-emptiness of CEFAs in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$ to the satisfiability of an LIA formula. The reduction relies on the following two observations. 
%The first observation follows from the fact that $\Lambda'_{x_i}$ for each $i \in [n]$ contains exactly one CEFA, and the second observation follows from the fact that the non-emptiness of exactly one CEFA w.r.t an LIA formula is entirely unrelated to the alphabet.

\begin{observation}\label{obs-cefa-output}
CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ iff there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true.
\end{observation}

Let $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$ be a CEFA and $\costset_\aut = \{\myvec{v} \mid \exists q, a, q'.\ (q, a, q', \myvec{v}) \in \delta\}$. 
Moreover, let $\cU_\aut = (Q, \costset_\aut, \delta', I, F)$ be an NFA over the alphabet $\costset_\aut$ that is obtained from $\aut$ by dropping the accepting condition and ignoring the characters, that is, $\delta'$ comprises tuples $(q, \myvec{v}, q')$ such that $(q, a, q', \myvec{v}) \in \delta$ for $a \in \Sigma$.
%Then $\cefaout(\aut) = \parikh(\NFA')$.
%

@@

\item there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true, 
\item 
there are $\eta_1 \in \parikh(\cU_{\cB_1}), \cdots, \eta_n \in \parikh(\cU_{\cB_n})$ such that 
\[
\bigwedge \limits_{i \in [n]} \alpha_i\left[ \sum \limits_{\myvec{v} \in \costset_{\cB_i}} \eta_i(\myvec{v}) \myvec{v} \big{/}R_{\cB_i}  \right] \wedge \varphi \left[\sum \limits_{\myvec{v} \in \costset_{\cB_1}} \eta_1(\myvec{v}) \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum \limits_{\myvec{v} \in \costset_{\cB_n}} \eta_n(\myvec{v}) \myvec{v} \big{/}R_{\cB_n} \right]
\]
is true. 
\end{itemize}

Therefore, to solve the non-emptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$, it is sufficient to compute the existential LIA formulas $\psi_{\cU_{\cB_1}}(\fZ_{\costset_{\cB_1}}), \cdots, \psi_{\cU_{\cB_n}}(\fZ_{\costset_{\cB_n}})$ to represent the Parikh images of $\cU_{\cB_1}$, $\cdots$, $\cU_{\cB_n}$ respectively, where $\fZ_{\costset_{\cB_i}} = \{\anivarz_{i,\myvec{v}} \mid \myvec{v} \in \costset_{\cB_i}\}$ for $i \in [n]$, and solve the satisfiability of the following existential LIA formula
\[
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cU_{\cB_i}}(\fZ_{\costset_{\cB_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cB_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right)\  \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cB_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cB_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\]
Intuitively, the integer variables $\anivarz_{i, \myvec{v}}$ represent the number of occurrences of $\myvec{v}$ in the strings accepted by $\cU_{\cB_i}$.

@@

\medskip
\noindent
\emph{Step 3 (Computing Parikh images).} For each $i \in [n]$, we compute an existential LIA formula $\psi_{\cC_i}(\fZ_{\costset_{\cC_i}})$ from $\cC_i$ to represent $\parikh(\cC_i)$. Then we solve the satisfiability of the following formula,
%
\[
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cC_i}(\fZ_{\costset_{\cC_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cC_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right) \ \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cC_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cC_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\]
%\qed

>>>>

\begin{definition}[$\cefadec$]
Let $x_1, \cdots, x_n$ be string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ be nonempty sets of CEFA over the alphabet $\Sigma$ with $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$ where the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{1, l_1}}, \cdots, R_{\aut_{n, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, moreover, let $\varphi$ be an LIA formula whose free variables are from $ \bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$. Then  the CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are said to be nonempty w.r.t. $\varphi$ if there are assignments $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v}_{i,j}$
such that $(\theta(x_i); \myvec{v}_{i,j}) \in \Lang(\aut_{i, j})$ and  $\varphi[(\myvec{v}_{i,j}/R_{\aut_{i,j}})]$ is true, for every $i \in [n], j \in [l_i]$.
\end{definition}

@@

At first, we show how to construct a CEFA from a regex where counting operators may occur. \hdh{Glushkov’s construction~\cite{Gluskov61} and its variants~\cite{redos_lenka,HS+23,GKM23} convert a regex to a $\epsilon$-free automaton. In contrast, Thompson's construction~\cite{Thompson1968} results in an automaton with $\epsilon$-transitions but the algorithm is simpler and straightforward. }Our approach is based on Thompson's construction~\cite{Thompson1968} \hdh{and we remove $\epsilon$-transitions on the fly (See union at Section \ref{sec:automaton} and construction of counting operators below)}. Let us start with register-representable regexes defined in the sequel.

@@

Let $e$ be a regex over $\Sigma$. Then an occurrence of counting operators in $e$, say $(e')^{\{m,n\}}$ (or $(e')^{\{m, \infty\}}$), is said to be \emph{register-representable} if $(e')^{\{m,n\}}$ (or $(e')^{\{m, \infty\}}$) is not in the scope of a Kleene star, another counting operator, complement, or language difference in $e$. We say that $e$ is \emph{register-representable} if and only if all occurrences of counting operators in $e$ are register-representable. For instance, $a^{\{2, 6\}} \cap a^{\{4, \infty\}}$ is register-representable, while $\overline{a^{\{2, 6\}}}$ and $(a^{\{2, 6\}})^{\{4,\infty\}}$ are not since $a^{\{2,6\}}$ is in the scope of complement and the counter operator $\{2, \infty\}$ respectively.

@@

For non-register-representable regexes, we first convert them into register-representable regexes by unfolding all the non-register-representable occurrences of counting operators. After that, we utilize the aforementioned procedure to construct CEFA. For instance, $(a^{\{2, 6\}} \concat b^*)^{\{2,\infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa+aaaa) \concat b^*)^{\{2, \infty\}}$. 
%$(a^{\{2, 6\}} \concat b^*) \setminus (a + b)^{\{4, \infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa + aaaa) \concat b^*) \setminus (a+b)(a+b)(a+b)(a+b)(a+b)^*$.

@@

In this section, we present a procedure to solve the $\cefadec$ problem: Suppose that $x_1, \cdots, x_n$ are mutually distinct string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty sets of CEFA over the same alphabet $\Sigma$ where $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots,$ $ \aut_{i, l_i}\}$ for every $i \in [n]$. Moreover, the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, and $\varphi$ is a LIA formula whose free variables are from $\bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$. 
The procedure comprises three steps.

@@

In the following steps, we reduce the non-emptiness of CEFAs in $\Lambda'_{x_1}, \cdots, $ $\Lambda'_{x_n}$ w.r.t. $\varphi$ to the satisfiability of an LIA formula. The reduction relies on the following two observations. 
%The first observation follows from the fact that $\Lambda'_{x_i}$ for each $i \in [n]$ contains exactly one CEFA, and the second observation follows from the fact that the non-emptiness of exactly one CEFA w.r.t an LIA formula is entirely unrelated to the alphabet.

\begin{observation}\label{obs-cefa-output}
CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ iff there are \hdh{outputs} $\myvec{v}_1 \in \cefaout(\cB_1), \cdots, \myvec{v}_n \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v}_1/R_{\cB_{1}}, \cdots, $ $\myvec{v}_n/R_{\cB_{n}}]$ is true.
\end{observation}

Let $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$ be a CEFA and $\costset_\aut = \{\myvec{v} \mid \exists q, a, q'.\ (q, a, q', \myvec{v}) $ $\in \delta\}$. 
Moreover, let $\cU_\aut = (Q, \costset_\aut, \delta', I, F)$ be an NFA over the alphabet $\costset_\aut$ that is obtained from $\aut$ by dropping the accepting condition and ignoring the characters, that is, $\delta'$ comprises tuples $(q, \myvec{v}, q')$ such that $(q, a, q', \myvec{v}) \in \delta$ for $a \in \Sigma$.
%Then $\cefaout(\aut) = \parikh(\NFA')$.
%

@@

\item there are \hdh{outputs} $\myvec{v}_1 \in \cefaout(\cB_1), \cdots, \myvec{v}_n \in \cefaout(\cB_n)$ such that \\$\varphi[\myvec{v}_1/R_{\cB_{1}}, \cdots, \myvec{v}_n/R_{\cB_{n}}]$ is true, 
\item 
there are $\eta_1 \in \parikh(\cU_{\cB_1}), \cdots, \eta_n \in \parikh(\cU_{\cB_n})$ such that 
\[
\bigwedge \limits_{i \in [n]} \alpha_i\left[ \sum \limits_{\myvec{v} \in \costset_{\cB_i}} \eta_i(\myvec{v}) \myvec{v} \big{/}R_{\cB_i}  \right] \wedge \varphi \left[\sum \limits_{\myvec{v} \in \costset_{\cB_1}} \eta_1(\myvec{v}) \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum \limits_{\myvec{v} \in \costset_{\cB_n}} \eta_n(\myvec{v}) \myvec{v} \big{/}R_{\cB_n} \right]
\]
is true. 
\end{itemize}

Therefore, to solve the non-emptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$, it is sufficient to compute the existential LIA formulas $\psi_{\cU_{\cB_1}}(\fZ_{\costset_{\cB_1}}), \cdots, $ $\psi_{\cU_{\cB_n}}(\fZ_{\costset_{\cB_n}})$ to represent the Parikh images of $\cU_{\cB_1}$, $\cdots$, $\cU_{\cB_n}$ respectively, where $\fZ_{\costset_{\cB_i}} = \{\anivarz_{i,\myvec{v}} \mid \myvec{v} \in \costset_{\cB_i}\}$ for $i \in [n]$, and solve the satisfiability of the following existential LIA formula
\[
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cU_{\cB_i}}(\fZ_{\costset_{\cB_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cB_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right)\  \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cB_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cB_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\]
Intuitively, the integer variables $\anivarz_{i, \myvec{v}}$ represent the number of occurrences of $\myvec{v}$ in the strings accepted by $\cU_{\cB_i}$.

@@

\medskip
\noindent
\emph{Step 3 (Computing Parikh images \hdh{and derive a solution}).} For each $i \in [n]$, we compute an existential LIA formula $\psi_{\cC_i}(\fZ_{\costset_{\cC_i}})$ from $\cC_i$ to represent $\parikh(\cC_i)$. \hdh{Subsequently, we can employ established SMT solvers to determine the satisfiability of the formula and derive a solution if it is satisfiable. The integer variable solutions are directly extracted from the partial model of the formula. The string variable solutions are obtained through indirect means, whose detailed methodology is discussed in Section \ref{sec-opt-sol-gen}. The final Parikh images are as follow :}
%
\[
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cC_i}(\fZ_{\costset_{\cC_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cC_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right) \ \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cC_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cC_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\]
%\qed

\hdh{
\subsection{Complexity Analysis of the $\reclsat$ Problem}
The time complexity of regular expression equivalence with complement has been established as nonelementary~\cite{complementary_complexity_1974}. This directly implies that the $\reclsat$ problem exhibits a nonelementary lower bound. We now turn our attention to analyzing the complexity of the $\reclsat$ problem in the absence of complement operations.

\begin{lemma}\label{lemma:regex2cefa}
    Register-representable regex without complement operation can be converted to CEFA in PSPACE.
\end{lemma}

\begin{proof}
    We prove this lemma through structural induction on regular expression $e$. For the base case—where the regular expression consists of a single character, $\epsilon$, or $\emptyset$—the lemma holds trivially.
    For the inductive cases, assuming the lemma holds for regular expressions $e_1$ and $e_2$:
    When $e = e_1\concat e_2$ or $e = e_1+e_2$, the concatenation or union operations described in Section \ref{sec:automaton} produce a CEFA of size $O(|e_1|^{k_1} + |e_2|^{k_2})=O(|e|^{\max(k_1, k_2)})$.
    For $e = e_1\cap e_2$, the intersection operation detailed in Section \ref{sec:automaton} yields a CEFA of size $O(|e_1|^{k_1}\cdot|e_2|^{k_2})=O(|e|^{k_1+k_2})$.
    In the case of $e=e_1^{\{m,n\}}$, our construction in Section \ref{subsec:regex2cefa} avoids direct unfolding of counting operators, resulting in a CEFA of size $O(|e_1|^{k_1})=O(|e|^{k_1})$.
    Finally, for $e=(e_1)$ and $e=e^*$, our application of Thompson's construction~\cite{Thompson1968} produces a CEFA of size $O(|e_1|^{k_1})=O(|e|^{k_1})$.
\end{proof}

\begin{theorem}\label{theorem_pspace}
    If all regexes are register-representable and complement operations are excluded, $\reclsat$ can be solved in PSPACE-complete.
\end{theorem}
\begin{proof}
    Based on Lemma \ref{lemma:regex2cefa}, our construction yields a CEFA of polynomial size for each register-representable regex without complement operation. Furthermore, for each string variable's length function, a CEFA of constant size is constructed in Section \ref{subsec:regex2cefa}.
    Consequently,  the construction outlined in Section \ref{subsec:regex2cefa} requires nondeterministic polynomial size. By Proposition \ref{prop-cefadec}, we can demonstrate that the procedure described in Section \ref{subsec:cefadec} likewise requires nondeterministic polynomial size. Therefore, the complexity of the complete decision procedure for solving the restricted $\reclsat$ problem is NPSPACE-complete, which yields PSPACE-complete by Savitch's theorem~\cite{complexity-book}.
\end{proof}

\begin{theorem}
    $\reclsat$ can be solved in EXPSPACE when complement operations are excluded.
\end{theorem}

\begin{proof}
    In Section \ref{subsec:regex2cefa}, unfolding non-register-representable occurrences of counting operators leads to an exponential growth in expression length. For each non-register-representable regular expression of length $l$, the unfolding process generates a register-representable regular expression of length $O(m^n\cdot l)=O((2^{|l|})^{|l|}\cdot l) = O(2^{|l|})$, where $m$ represents the maximum counting bound and $n$ denotes the maximum depth of nested counting operators. By Theorem \ref{theorem_pspace}, the complexity of the complete decision procedure for solving the $\reclsat$ problem without complement operations is EXPSPACE.
\end{proof}
}
